Artificial selection for microbial collective composition can succeed or fail depending on the initial and target values

Version 1

Abstract

Microbial collectives, capable of functions beyond the reach of individual populations, can be enhanced through artificial selection. However, this process presents unique challenges. Here, we explore the ‘waterfall’ phenomenon, a metaphor describing how the success in achieving a desired genotype or species composition in microbial collectives can depend on both the target characteristics and initial conditions. We focus on collectives comprising fast-growing (F) and slow-growing (S) types, aiming to achieve specific S frequencies. Through simulations and analytical calculations, we show that intermediate target S frequencies might be elusive, akin to maintaining a raft’s position within a waterfall, rather than above or below it. This challenge arises because intra-collective selection, favoring F during growth, is the strongest at intermediate S frequencies, which can overpower counteracting inter-collective selection effects. Achieving low target S frequencies is consistently possible as expected, but high target S frequencies require an initially high S frequency — similar to a raft that can descend but not ascend a waterfall. The range of attainable target frequencies is significantly influenced by the initial population size of the collectives, while the number of collectives under selection plays a less critical role. In scenarios involving more than two types, the evolutionary trajectory must navigate entirely away from the metaphorical ‘waterfall drop.’ Our findings illustrate that the strength of intra-collective evolution is frequency-dependent, with implications in experimental planning.

Introduction

Microbial collectives can carry out functions that arise from interactions among member species. These functions, such as waste degradation (1, 2), probiotics (3), and vitamin production (4), can be useful for human health and biotechnology. To improve collective functions, one can perform artificial selection (directed evolution) on collectives: Low-density “Newborn” collectives are allowed to “mature” during which cells proliferate and possibly mutate, and community function develops. “Adult” collectives with high functions are then chosen to reproduce, each seeding multiple offspring Newborns. Artificial selection of collectives have been attempted both in experiments (5–19) and in simulations (20–30), often with unimpressive outcomes.

One of the major challenges in selecting collectives is to ensure the inheritance of a collective function (31, 32). Inheritance can be compromised when genotype and species compositions change from a parent collective to offspring collectives. During maturation of a collective, genotype compositions within each species can be altered by intra-collective evolution, while species compositions can be shifted by ecological interactions. Furthermore, during reproduction of a collective, genotype and species compositions of offspring can vary stochastically from those of the parent.

Here, we consider the selection of collectives of two types with different growth rates to achieve a taget composition in the Adult collective. Earlier work has demonstrated that nearly any target species composition can be achieved when selecting communities of two competing species with unequal growth rates (24, 33), so long as the shared resource is depleted during collective maturation (24). In this case, both species initially evolved to grow faster, and the slower-growing species was preserved due to stochastic fluctuations in species composition during collective reproduction. Eventually, both species evolved to grow sufficiently fast to deplete the shared resource during collective maturation, and evolution in competition coefficients then acted to stabilize species ratio to the target value (24).

Here, we mathematically examine the selection of target type compositions in two-type collectives when nutrient is always in excess. This allows us to analytically examine the evolutionary tipping point between intra-collective and inter-collective selection. We show that this tipping point creates a “waterfall” effect which restricts not only which target compositions are achievable, but also the initial composition required to achieve the target. We also investigate how the range of achievable target composition is affected by the population size in Newborns and the total number of collectives under selection. Finally, we show that the waterfall phenomenon extends to systems with more than 2 types.

Results and Discussions

We start with selecting for a target composition in collectives of fast-growing F and slow-growing S types. We show that intermediate F frequencies or equivalently, intermediate S frequencies, are the hardest targets to achieve. We then show that similar conclusions hold when selecting for a target composition in collectives of more than two types. We mainly describe in terms of F frequency f = F/(S + F), which is related to S frequency s = 1 − f .

A selection cycle

A selection cycle (Fig. 1) starts with a total of g Newborn collectives. We consider two genotypes - slow-growing type (S) and fast-growing type (F), with S mutating to F at a rate μ. At the beginning of cycle k (t = 0), a collective has a fixed total cell number where and denote the numbers of S and F cells in collective i (1 ≤ i ≤ g) at time t (0 ≤ t ≤ τ) of cycle k. The average F frequency among the g Newborn collectives is , such that the initial F cell number in each Newborn is drawn from the binomial distribution Binom .

Schematic for artificial selection on collectives.

Each selection cycle begins with a total of g Newborn collectives (black open circles), each with N0 total cells of slow-growing type (S, red-colored dots) and fast-growing type (F, blue-colored dots). During maturation (over time τ), S and F cells divide at rates r and r + ω (ω > 0), respectively. S mutates to F at rate μ. In the selection stage, the Adult collective with F frequency f closest to the target composition is chosen to reproduce g Newborns for the next cycle. Newborns are sampled from the chosen Adult (yellow star) with N0 cells each Newborn. The selection cycle is then repeated until the F frequency reaches a steady state, which may or may not be the target composition. To denote a quantity x of i-th collective in cycle k at time t (0 ≤ t ≤ τ), we use notation where x ∈ {S, F, s, f }. Note that time t = 0 is for Newborns and t = τ is for Adults.

Collectives are allowed to grow for time τ (‘Maturation’ in Fig. 1). During maturation, S and F grow at rates r and r + ω (ω > 0), respectively. If maturation time τ is too small, matured collectives (“Adults”) do not have enough cells to reproduce g Newborn collectives with N0 cells. On the other hand, if maturation time τ is too long, fast-growing F will take over. Hence we set the maturation time τ = ln(g + 1)/r, which guarantees sufficient cells to produce g Newborn collectives from a single Adult collective.

At the end of a cycle, the Adult with the highest function is chosen to reproduce g Newborn collectives each with N0 cells (‘Selection’ and ‘Reproduction’ in Fig. 1). This allows us to employ the simplest version of the extreme value theory, and in our case, choosing top 1 outperforms a less stringent “choosing top 5%” selection regime. Collective function is dictated by the F frequency f, and is maximized at the target value . Thus among all Adult collectives, the chosen Adult has F frequency f ∗ closest to the target value . For the next cycle, the number of F cells in Newborns follow a binomial distribution with the mean value N0f ∗. By repeating the selection cycle, we aim to achieve the target composition .

The success of collective selection is constrained by the target composition, and sometimes also by the initial composition

Since intra-collective selection favors F, we expect that a higher target (a lower target ŝ) is easier to achieve. Note that once a target frequency is achieved, inter-collective selection is always required to maintain that frequency due to the fitness difference between the two types.

We fixed N0,the total population size of a Newborn, to 1000, and obtained selection dynamics for various initial and target F frequencies by implementing stochastic simulations (Sec. 1 in Supplementary Information). If the target value is high (e.g., Figure 2a magenta), selection is successful in the sense that the absolute error d between the target frequency and the selected frequency averaged among independent simulations ⟨f ∗⟩ is smaller than 0.05 (i.e computed absolute errors are shown in Supplementary Information Fig. S3). F frequency of the chosen collective eventually converges to the target value and stays around it, regardless of the initial frequency. In contrast, without collective-level selection (e.g. choosing a random collective to reproduce), F frequency increases until F reaches fixation (Supplementary information Fig. S4b).

Initial and target compositions determine the success of artificial selection on collectives.

(a-c) Mutant frequency of the selected Adult collective (f ∗) over cycles. The target frequency is marked as a dotted line, while f H (black solid line) and f L (black dashed line) delineate boundary values that define the region of successful selection. a A high target F frequency ( ; magenta dotted line) can be achieved from any initial frequency (black dots). b An intermediate target frequency ( ; green dotted line) is never achievable, as all initial conditions converge to near f H . c A low target freqeuncy is acheiveable, but only from initial frequencies below f L. For initial frequencies at f L, stochastic outcomes (grey curves) are observed: while some replicates reached the target frequency, other reached f H. For parameters, we used wildtype growth rate r = 0.5, F growth advantage ω = 0.03, mutation rate μ = 0.0001, maturation time τ ≈ 4.8, and N0 = 1000. The number of collective g = 10. Each black line is averaged from independent 300 realizations. d Two accessible regions (gold). Either high (region 2) or low starting from low initial f (region 1) can be achieved. We theoretically predict (by numerically integrating Eq. 1) the boundaries of success regions, f H (black solid line) and f L (black dashed lines), which agree with simulation results (gold regions). e Example trajectories from initial compositions (black dots) to the target compositions (dashed lines). The gold areas indicate the region of initial frequencies where the target frequency can be achieved. f The tension between intra-collective selection and inter-collective selection creates a “waterfall” phenomenon. See the main text for details.

In contrast, an intermediate target frequency (e.g. Fig. 2b green) is never achievable. High initial F frequencies (e.g. 0.6 and 0.95) decline toward the target, but stabilize at an “higher-threshold” (f H ∼ 0.7, solid line segment in Fig. 2a-c) above the target. Low initial F frequencies (e.g. 0) increase toward the target, but then overshoot and stabilize at the f H value.

If the target frequency is low (e.g. Fig. 2c red line), artificial selection succeeds when the initial frequency is below a “lower-threshold” (f L, dashed line segment in Fig. 2a-c). Initial F frequencies above f L (e.g. 0.45 and 0.95) converge to f H instead. Initial F frequencies near f L display stochastic trajectories, converging to either f H or the target.

In summary, two regions of target frequencies are “accessible” in the sense that a target value in the region can be maintained once reached; gold in Fig. 2d, e): (1) target frequencies above and (2) target frequencies below and starting at an average frequency below .

Intra-collective evolution is the fastest at intermediate F frequencies, creating the “waterfall” phenomenon

To understand what gives rise to the two accessible regions, we calculated how the distribution of F frequency f might change after one cycle. We consider the case where , the F frequency of the selected Adult at cycle k, is above the target value . This case is particularly challenging because intra-collective evolution favours fast-growing F and thus will further increase f away from the target. From , Newborns of cycle k + 1 will have f fluctuating around , and after they mature,in F frequency over one cycle, the minimum f is selected . If the selected composition at cycle k + 1 can be reduced compared to that of cycle k (i.e. ), the system can evolve to the lower target value.

Mathematically, the conditional probability distribution Ψ of (the F frequency of the selected Adult at cycle k + 1 being f) given the selected at cycle k is Equation 1 can be broken down into two parts: (i) describes that any one of the g collectives must achieve f, where denotes the probability of an Adult (t = τ) in cycle k + 1 achieving f given . (ii) describes that the remaining g − 1 collectives must all achieve a frequency above f (and thus not selected). is calculated by integrating across ζ the term (the probability density of fk+1,0 of a Newborn in cycle k + 1 achieving ζ given ) multiplied by the term (the probability density of an Adult achieving f′ given Newborn’s frequency ζ, see Methods).

When we approximate the Adults’ f in cycle k + 1 as Guassian with mean and variance , we can obtain an analytical solution of the change in f over one cycle - the difference between the median value of the conditional probability distribution Ψ and the selected frequency of (Sec. 2 in Supplementary Information): where and e is Euler’s number (2.71828…), and Φ−1(…) is an inverse cumulative function of standard normal distribution. For those unfamiliar with Φ−1(…), y = Φ−1(x) (where 0 ≤ x ≤ 1) is the number such that the probability of a standard normal random variable being less than or equal to y is x (e.g. 0 = Φ−1(0.5)). Eq. [2] contains all experimental parameters, including g (the total number of collectives under selection), τ (maturation time), μ (mutation rate), r (growth rate of the slow-growing S type), and ω (growth advantage of F over S). Equation [2] has the same shape as results from numerically integrating Eq. [1] and from stochastic simulations (Fig. 3a).

Intra-collective selection and inter-collective selection jointly set the boundaries for selection success.

a The change in F frequency over one cycle. When is sufficiently low or high, inter-collective selection can lower the F frequency to below . The points where Median are denoted as f L and f H, corresponding to the boundaries in Fig. 2. b The distributions of frequency differences obtained by 1000 numerical simulations. The cyan, purple, and black box plots indicate the changes in F frequency after intra-collective selection during maturation, after inter-collective selection, and over one selection cycle, respectively. The box ranges from 25% to 75% of the distribution, and the median is indicated by a line across the box. The upper and lower whiskers indicate maximum and minimum values of the distribution. ***: P < 0.001 in an unpaired t-test.

F frequency can be reduced if the selected is low or high, but not if it is intermediate between f L and f H (Fig. 3a). Increase in f during maturation is the most drastic when Newborn f is around 0.5 (Fig. 3b blue). This is intuitive: when Newborn f is low, the increase will be minor; when Newborn f is high, the fitness advantage of F over the population average is small and hence the increase is also minor. Thus, when Newborn F frequency is intermediate, inter-collective selection may not be able to counter intra-collective selection (Fig. 3b).

Not surprisingly, similar conclusions are derived where S and F are slow-growing and fast-growing species which cannot be converted through mutations (Sec. 3 and Fig. S7 in Supplementary Infomation).

Overall, inter-collective selection is akin to a raftman, rowing the raft to a target, while intra-collective selection is akin to a waterfall. This metaphore is best understood in terms of S frequency s = 1 − f . Then, the lower-threshold f L corresponds to higher-threshold in sH = 1 − f L, while higher-threshold f H corresponds to lower-threshold in sL = 1 − f H . Intra-collective selection is akin to a waterfall, driving the S frequency s from high to low (Fig. 2f). Intra-collective selection acts the strongest when s is intermediate (sL < s < sH), similar to the vertical drop of the fall. Intra-collective selection acts weakly at high (> sH) or low (< sL) s, similar to the gentle sloped upper and lower pools of the fall (regions 1 and 2 of Fig. 2d, f and Fig. 3a). Thus, an intermediate target frequency can be impossible to achieve - a raft starting from the upper pool will be flushed down to sL (f H), while a raft starting from the lower pool cannot go beyond sL (f H). In contrast, a low target S frequency (in the lower pool) is always achievable. Finally, a high target S frequency (in the upper pool) can only be achieved if starting from the upper pool (as the raft can not jump to the upper pool if starting from below).

Manipulating experimental setups to expand the achievable target region

Variation among collectives is a key element for successful selection. Here, variation among collectives originates from the stochastic nature of growth and mutation during collective maturation, and of sampling effects during collective reproduction. From Eq. [4], the variance of f at time τ scales with . Indeed, as Newborn size N0 declines, the region of acheivable of target frequency expands (larger gold area in Fig. 4a). If the Newborn size N0 is smaller than ∼700, any target frequency can be reached when other parameters are fixed.

Expanding the success region for artificial collective selection.

a Reducing the population size in Newborn N0 expands the region of success. In gold area, the probability that becomes smaller than in a cycle is more than 50%. We used g = 10 and τ ≈ 4.8. Figures 2-3 correspond to N0 = 1000. b Increasing the total number of collectives g also expands the region of success, although only slightly. We used a fixed Newborn size N0 = 1000. The maturation time τ (τ = log(100)/r) is set to be long enough so that an Adult can generate at least 100 Newborns.

The number of collectives g also affects the selection outcomes. With more collectives, the chance to find a f closer to the target is more likely. Thus, a larger number of collectives broadens the region of success (Fig. 4b). However, the effect of g is not dramatic. To see why, we note that the only place that g appears is Eq. [2] in the form of . When g becomes large, Φ−1(1/g) is asymptotically expressed as (Sec. 2 in Supplementary Information)(34). Thus, φ(g) does not change dramatically as g varies.

The waterfall phenomenon in a higher dimension

In contrast to the above two-type case where the evolutionary trajectory travels along a one-dimensional line from the initial frequency to the final frequency, in a three-type case the trajectory travels in a two-dimensional plane. To examine the waterfall effect in a higher dimension, we investigate a three-type system where a faster-growing type (FF) grows faster than the fast-growing type (F) which grows faster than the slow-growing type (S) (Fig. 5a and Sec. 6 in Supplementary Information). In this system, a target type composition can be achieved if inter-collective selection reduces the frequencies of F as well as FF (accessible region, gold in Fig. 5b). Strikingly, a target composition in an accessible region may not be achievable even when the initial composition is within the same region: once the composition escapes the accessible region, the trajectory cannot return back to the accessible region (Fig. 5biii, the leftmost initial condition). However, if the initial position is closer to the target in the accessible region, the target becomes achievable (Fig. 5biii, initial condition near the bottom).

In higher dimensions, the success of artificial selection requires the entire evolutionary trajectory remaining in the accessible region.

a During collective maturation, a slow-growing type (S) (with growth rate r; dark red) can mutate to a fast-growing type (F) (with growth rate r + ω; blue), which can mutate further into a faster-growing type (FF) (with growth rate r + 2ω; purple). Here, the rates of both mutational steps are μ, and ω > 0. b Evolutionary trajectories from various initial compositions (open circles) to various targets. Intra-collective evolution favors FF over F (vertical blue arrow) over S (horizontal blue arrow). The accessible regions are marked gold (see Sec. 1 in Supplementary Information). We obtain final compositions starting from several initial compositions while aiming for different target compositions in i, ii, and iii. The evolutionary trajectories are shown in dots with color gradients from the initial to final time. (i) A target composition with a high FF frequency is always achievable. (ii) A target composition with intermediate FF frequency is never achievable. (iii) A target composition with low FF frequency is achievable only if starting from an appropriate initial composition such that the entire trajectory never meanders away from the accessible region. The figures are drawn using mpltern package (35).

In conclusion, we have investigated the evolutionary trajectories of type composition in collectives under selection, which are governed by intra-collective selection (which favors fast-growing types) and inter-collective selection (which, in our case, strives to counter fast-growing types). Intra-collective selection has the strongest effect at intermediate frequencies of faster-growing types, potentially creating an inaccessible region analogous to the vertical drop of a waterfall. High and low target frequencies are both accessible, analogous to the lower- and the upper-pools of a waterfall, respectively. A less challenging target (low ŝ)is achievable from any initial position. In contrast, a more challenging target (high ŝ) is only achievable if the entire trajectory is contained within the region, similar to a raft striving to reach a point in the upper-pool must start at and remain in the upper pool. Our work suggests that the strength of intra-collective selection is not constant, and that strategically choosing an appropriate starting point can be essential for successful collective selection.

Materials and Methods

Stochastic simulations

A selection cycle is composed of three steps: maturation, selection, and reproduction. At the beginning of the cycle k, a collective i has slow-growing cells and fast-growing cells. At the first cycle, mean F frequency of collectives are set to be is sampled from the binomial distribution with mean . Then, S cells are in the collective i. In the maturation step, we calculate and by using stochastic simulation. We can simulate the division and mutation of each individual cell stochastically by using tau-leaping algorithm (36, 37)(see Fig. S4a and b in Supplementray Information). However, individual-based simulation requires a long computing time. Instead, we randomly sample and from the probability distribution . To obtain , we solve the master equation which describes a time evolution of probability distribution, under the random processes (see Sec. 1 in Supplementary Information). We assumed that follows the Gaussian distribution. The mean and variance of the distribution are numerically obtained from ordinary differential equations which is derived from the master equation (see Sec. 1 in Supplementary Information). We choose the collective with the closest frequency to the target . The selected collective generates g Newborns. The number of F cells is sampled from the binomial distribution with the mean of . We start a new cycle with those Newborn collectives. Then, the number of S cells in a collective i is

Analytical approach to the conditional probability

The conditional probability distribution Ψ that we observe at a given is calculated by the following procedure. Given the selected collective in cycle k with , the collective-level reproduction proceeds by sampling g Newborn collectives with N0 cells in cycle k +1. Each Newborn collective contains certain F numbers at the beginning of the cycle k + 1, which can be mapped into with the constraint of N0 cells. If the number of cells in the selected collective is large enough, the joint conditional distribution function is well described by the product of g independent and identical Gaussian distribution 𝒩(μ, σ2). So we consider the frequencies of g Newborn collectives as g identical copies of the Gaussian random variable fk+1,0. The mean and variance of fk+1,0 are given by and . Then, the conditional probability distribution function of fk+1,0 being ζ is given by After the reproduction step, the Newborn collectives grow for time τ . The frequency is changed from the given frequency ζ to f by division and mutation processes. We assume that the frequency f of an Adult is also approximated by a Gaussian random variable . The mean and variance are calculated by using means and variances of S and F (see Sec. 2 in Supplementary Information.) Since and also depend on ζ, the conditional probability distribution function of fk+1,τ being f is given by The conditional probability distribution of an Adult collective in cycle k + 1 (fk+1,τ) to have frequency f at a given is calculated by producting two Gaussian distribution functions and integrating overall ζ values, which is given by Since we select the minimum frequency among g identical copies of fk+1,τ, the conditional probability distribution function of follows a minimum value distribution, which is given in Eq. [1]. Here, For the case of , the selected frequency is the minimum frequency So we have by replace to

We assume that the conditional probability distribution in Eq. [7] follows a normal distribution, whose mean and variances are describe by Eq. [3] and Eq. [4]. Then, the extreme value theory (38) estimates the median of the selected Adult by The F frequency difference in Eq. [2] is obtained by substracting from Eq. [8].

eLife Assessment

This important study of artificial selection in microbial communities shows that the possibility of selecting a desired fraction of slow and fast-growing types is impacted by their initial fractions. The evidence, which relies on mathematical analysis and simulations of a stochastic model, is convincing. It highlights the tension between selection at the strain and the community level. This study should be of interest to researchers interested in ecology, both theoretical and experimental.

Reviewer #1 (Public Review):

Summary:

The authors demonstrate with a simple stochastic model that the initial composition of the community is important in achieving a target frequency during the artificial selection of a community.

Strengths:

To my knowledge, the intra-collective selection during artificial selection has not been seriously theoretically considered. However, in many cases, the species dynamics during the incubation of each selection cycle are important and relevant to the outcome of the artificial selection experiment. Stochasticity from birth and death (demographic stochasticity) plays a big role in these species' abundance dynamics. This work uses a simple framework to tackle this idea meticulously.

This work may or may not be related to hysteresis (path dependency). If this is true, maybe it would be nice to have a discussion paragraph talking about how this may be the case. Then, this work would even attract the interest of people studying dynamic systems.

Weaknesses:

(1) Connecting structure and function

In typical artificial selection literature, most of them select the community based on collective function. Here in this paper, the authors are selecting a target composition. Although there is a schematic cartoon illustrating the relationship between collective function (y-axis) and the community composition in the main Figure 1, there is no explicit explanation or justification of what may be the origin of this relationship. I think giving the readers a naïve idea about how this structure-function relationship arises in the introduction section would help. This is because the conclusion of this paper is that the intra-collective selection makes it hard to artificially select a community that has an intermediate frequency of f (or s). If there is really evidence or theoretical derivation from this framework that indeed the highest function comes from the intermediate frequency of f, then the impact of this paper would increase because the conclusions of this stochastic model could allude to the reasons for the prevalent failures of artificial selection in literature.

(2) Explain intra-collective and inter-collective selection better for readers.

The abstract, the introduction, and the result section use these terms or intra-collective and inter-collective selection without much explanation. A clear definition in the beginning would help the audience grasp the importance of this paper, because these concepts are at the core of this work.

(3) Achievable target frequency strongly depending on the degree of demographic stochasticity.

I would expect that the experimentalists would find these results interesting and would want to consider these results during their artificial selection experiments. The main Figure 4 indicates that the Newborn size N0 is a very important factor to consider during the artificial selection experiment. This would be equivalent to how much bottleneck is imposed on the artificial selection process in every iteration step (i.e., the ratio of serial dilution experiment). However, with a low population size, all target frequencies can be achieved, and therefore in these regimes, the initial frequency now does not matter much. It would be great for the authors to provide what the N0 parameter actually means during the artificial selection experiments. Maybe relative to some other parameter in the model. I know this could be very hard. But without this, the main result of this paper (initial frequency matters) cannot be taken advantage of by the experimentalists.

(4) Consideration of environmental stochasticity.

The success (gold area of Figure 2d) in this framework mainly depends on the size of the demographic stochasticity (birth-only model) during the intra-collective selection. However, during experiments, a lot of environmental stochasticity appears to be occurring during artificial selection. This may be out of the scope of this study. But it would definitely be exciting to see how much environmental stochasticity relative to the demographic stochasticity (variation in the Gaussian distribution of F and S) matters in succeeding in achieving the target composition from artificial selection.

(5) Assumption about mutation rates

If setting the mutation rates to zero does not change the result of the simulations and the conclusion, what is the purpose of having the mutation rates \mu? Also, is the unidirectional (S -> F -> FF) mutation realistic? I didn't quite understand how the mutations could fit into the story of this paper.

(6) Minor points

In Figure 3b, it is not clear to me how the frequency difference for the Intra-collective and the Inter-collective selection is computed.

In Figure 5b, the gold region (success) near the FF is not visible. Maybe increase the size of the figure or have an inset for zoom-in. Why is the region not as big as the bottom gold region?

Reviewer #2 (Public Review):

The authors provide an analytical framework to model the artificial selection of the composition of communities comprised of strains growing at different rates. Their approach takes into account the competition between the targeted selection at the level of the meta-community and the selection that automatically favors fast-growing cells within each replicate community. Their main finding is a tipping point or path-dependence effect, whereby compositions dominated by slow-growing types can only be reached by community-level selection if the community does not start and never crosses into a range of compositions dominated by fast growers during the dynamics.

These results seem to us both technically correct and interesting. We commend the authors on their efforts to make their work reproducible even when it comes to calculations via extensive appendices, though perhaps a table of contents and a short description of these appendices at the start of SI would help navigate them.

The main limitation in the current form of the article is that it could clarify how its assumptions and findings differ from and improve upon the rest of the literature:

- Many studies discuss the interplay between community-level evolution and species- or strain-level evolution. But "evolution" can be a mix of various forces, including selection, drift/randomness, and mutation/innovation.

- This work's specificity is that it focuses strictly on constant community-level selection versus constant strain-level selection, all other forces being negligible (neither stochasticity nor innovation/mutation matter at either level, as we try to clarify now).

- Regarding constant community-level selection, it is only briefly noted that "once a target frequency is achieved, inter-collective selection is always required to maintain that frequency due to the fitness difference between the two types" [pg. 3 {section sign}2]. In other words, action from the selector is required indefinitely to maintain the community in the desired state. This assumption is found in a fraction of the literature, but is still worth clarifying from the start as it can inform the practical applicability of the results.

- More importantly, strain-level evolution also boils down here to pure selection with a constant target, which is less usual in the relevant literature. Here, (1) drift from limited population sizes is very small, with no meaningful counterbalancing of selection, (2) pure exponential regime with constant fitness, no interactions, no density- or frequency-dependence, (3) there is no innovation in the sense that available types are unchanging through time (no evolution of traits such as growth rate or interactions) and (4) all the results presented seem unchanged when mutation rate mu = 0 (as noted in Appendix III), meaning that the conclusions are not "about" mutation in any meaningful way.

- Furthermore, the choice of mutation mechanism is peculiar, as it happens only from slow to fast grower: more commonly, one assumes random non-directional mutations, rather than purely directional ones from less fit to fitter (which is more of a "Lamarckian" idea). Given that mutation does not seem to matter here, this choice might create unnecessary opposition from some readers or could be considered as just one possibility among others.

It would be helpful to have all these points stated clearly so that it becomes easy to see where this article stands in an abundant literature and contributes to our understanding of multi-level evolution, and why it may have different conclusions or focus than others tackling very similar questions.

Finally, a microbial context is given to the study, but the assumptions and results are in no way truly tied to that context, so it should be clear that this is just for flavor.

Reviewer #3 (Public Review):

The authors address the process of community evolution under collective-level selection for a prescribed community composition. They mostly consider communities composed of two types that reproduce at different rates, and that can mutate one into the other. Due to such differences in 'fitness' and to the absence of density dependence, within-collective selection is expected to always favour the fastest grower, but the collective-level selection can oppose this tendency, to a certain extent at least. By approximating the stochastic within-generation dynamics and solving it analytically, the authors show that not only high frequencies of fast growers can be reproducibly achieved, aligned with their fitness advantage. Small target frequencies can also be maintained, provided that the initial proportion of fast growers is sufficiently small. In this regime, similar to the 'stochastic corrector' model, variation upon which selection acts is maintained by a combination of demographic stochasticity and of sampling at reproduction. These two regions of achievable target compositions are separated by a gap, encompassing intermediate frequencies that are only achievable when the bottleneck size is small enough or the number of communities is (disproportionately) larger.

A similar conclusion, that stochastic fluctuations can maintain the system over evolutionary time far from the prevalence of the faster-growing type, is then confirmed by analyzing a three-species community, suggesting that the qualitative conclusions of this study are generalizable to more complex communities.

I expect that these results will be of broad interest to the community of researchers who strive to improve community-level selection, but are often limited to numerical explorations, with prohibitive costs for a full characterization of the parameter space of such embedded populations. The realization that not all target collective functions can be as easily achieved and that they should be adapted to the initial conditions and the selection protocol is also a sobering message for designing concrete applications.

A major strength of this work is that the qualitative behaviour of the system is captured by an analytically solvable approximation so that the extent of the 'forbidden region' can be directly and generically related to the parameters of the selection protocol.

I however found the description of the results too succinct and I think that more could be done to unpack the mathematical results in a way that is understandable to a broader audience. Moreover, the phenomenon the authors characterize is of purely ecological nature. Here, mutations of the growth rate are, in my understanding, neither necessary (non-trivial equilibria can be maintained also when \mu =0) nor sufficient (community-level selection is necessary to keep the system far from the absorbing state) for the phenomenon described. Calling this dynamics community evolution reflects a widespread ambiguity, and is not ascribable just to this work. I find that here the authors have the opportunity to make their message clearer by focusing on the case where the 'mutation' rate \mu vanishes (Equations 39 & 40 of the SI) - which is more easily interpretable, at least in some limits - while they may leave the more general equations 3 & 4 in the SI. Combined with an analysis of the deterministic equations, that capture the possibility of maintaining high frequencies of fast growers, the authors could elucidate the dynamics that are induced by the presence of a second level of selection, and speculate on what would be the result of real open-ended evolution (not encompassed by the simple 'switch mutations' generally considered in evolutionary game theory), for instance discussing the invasibility (or not) of mutant types with slightly different growth rates.

The single most important model hypothesis that I would have liked to be discussed further is that the two types do not interact. Species interactions are not only essential to achieve inheritance of composition in the course of evolution but are generally expected to play a key role even on ecological time scales. I hope the authors plan to look at this in future work.

Author response:

The following is the authors’ response to the original reviews.

Common comments

(1) Significance of zero mutation rate

Reviewers asked why we included mutation rate even though setting mutation rate to zero doesn’t change results. We think that including non-zero mutation rate makes our results more generalisable, and thus is a strength rather than weakness. To better motivate this choice, we have added a sentence to the beginning of Results:

(2) Writing the mu=0 case first

Reviewers suggested that we should first focus on the mu=0 case, and then generalize the result. The suggestions are certainly good. However, given the large amount of work involved in a re-organization, we have decided to adhere to our current narrative. However, we now only include equations where mu=0 in the main text, and have moved the case of nonzero mutation rate to Supplementary Information.

(3) Making equations more accessible

We have taken three steps to make equations more readable.

● Equations in the main text correspond to the case of zero-mutation rate.

● The original section on equation derivation is now in a box in the main text so that readers have the choice of skipping it but interested readers can still get a gist of where equations came from.

● We have provided a much more detailed interpretation of the equation (see page 10).

(4) Validity of the Gaussian approximation

Reviewers raised concerns about the validity of Gaussian approximation on F frequency𝑓(𝜏). The fact that our calculations closely match simulations suggest that this approximation is reasonable. Still, we added a discussion about the validity of this approximation in Box 1.

We also added to SI with various cases of initial S and F sizes. This figure shows that when either initial S or initial F is small, the distribution of𝑓(𝜏) is not normal. However, if initial S and F are both on the order of hundreds, then the distribution of 𝑓(𝜏) is approximately Gaussian.

Public Reviews:

Summary:

The authors demonstrate with a simple stochastic model that the initial composition of the community is important in achieving a target frequency during the artificial selection of a community.

Strengths:

To my knowledge, the intra-collective selection during artificial selection has not been seriously theoretically considered. However, in many cases, the species dynamics during the incubation of each selection cycle are important and relevant to the outcome of the artificial selection experiment. Stochasticity from birth and death (demographic stochasticity) plays a big role in these species' abundance dynamics. This work uses a simple framework to tackle this idea meticulously.

This work may or may not be hysteresis (path dependency). If this is true, maybe it would be nice to have a discussion paragraph talking about how this may be the case. Then, this work would even attract the interest of people studying dynamic systems.

We have added this clarification in the main text:

“Note that here, selection outcome is path-dependent in the sense of being sensitive to initial conditions. This phenomenon is distinct from hysteresis where path-dependence results from whether a tuning parameter is increased or decreased.

Weaknesses:

(1) Connecting structure and function

In typical artificial selection literature, most of them select the community based on collective function. Here in this paper, the authors are selecting a target composition. Although there is a schematic cartoon illustrating the relationship between collective function (y-axis) and the community composition in the main Figure 1, there is no explicit explanation or justification of what may be the origin of this relationship. I think giving the readers a naïve idea about how this structure-function relationship arises in the introduction section would help. This is because the conclusion of this paper is that the intra-collective selection makes it hard to artificially select a community that has an intermediate frequency of f (or s). If there is really evidence or theoretical derivation from this framework that indeed the highest function comes from the intermediate frequency of f, then the impact of this paper would increase because the conclusions of this stochastic model could allude to the reasons for the prevalent failures of artificial selection in literature.

We have added this to introduction: “This is a common quest: whenever a collective function depends on both populations, collective function is maximised, by definition, at an intermediate frequency (e.g. too little of either population will hamper function [23]).”

(2) Explain intra-collective and inter-collective selection better for readers.

The abstract, the introduction, and the result section use these terms or intra-collective and inter-collective selection without much explanation. For the wide readership of eLife, a clear definition in the beginning would help the audience grasp the importance of this paper, because these concepts are at the core of this work.

This is a great point. We have added in Abstract:

“Such collective selection is dictated by two opposing forces: during collective maturation, intra-collective selection acts like a waterfall, relentlessly driving the S-frequency to lower values, while during collective reproduction, inter-collective selection resembles a rafter striving to reach the target frequency. Due to this model structure, maintaining a target frequency requires the continued action of inter-collective selection.”

and in Introduction

“A selection cycle consists of three stages (Fig. 1). During collective maturation, intra-collective selection favors fast-growing individuals within a collective. At the end of maturation, inter-collective selection acts on collectives and favors those achieving the target composition. Finally during collective reproduction, offspring collectives sample stochastically from the parents, a process dominated by genetic drift.”

(3) Achievable target frequency strongly depending on the degree of demographic stochasticity.

I would expect that the experimentalists would find these results interesting and would want to consider these results during their artificial selection experiments. The main Figure 4 indicates that the Newborn size N0 is a very important factor to consider during the artificial selection experiment. This would be equivalent to how much bottleneck is imposed on the artificial selection process in every iteration step (i.e., the ratio of serial dilution experiment). However, with a low population size, all target frequencies can be achieved, and therefore in these regimes, the initial frequency now does not matter much. It would be great for the authors to provide what the N0 parameter actually means during the artificial selection experiments. Maybe relative to some other parameter in the model. I know this could be very hard. But without this, the main result of this paper (initial frequency matters) cannot be taken advantage of by the experimentalists.

We have added an analytical approximation for N0˘, the Newborn size below which all target frequencies can be achieved in SI.

Also, we have added lines indicating N0˘ in Fig4a.

(4) Consideration of environmental stochasticity.

The success (gold area of Figure 2d) in this framework mainly depends on the size of the demographic stochasticity (birth-only model) during the intra-collective selection. However, during experiments, a lot of environmental stochasticity appears to be occurring during artificial selection. This may be out of the scope of this study. But it would definitely be exciting to see how much environmental stochasticity relative to the demographic stochasticity (variation in the Gaussian distribution of F and S) matters in succeeding in achieving the target composition from artificial selection.

You are correct that our work considers only demographic stochasticity.

Indeed, considering other types of stochasticity will be an exciting future research direction. We added in the main text:

“Overall our model considers mutational stochasticity, as well as demographic stochasticity in terms of stochastic birth and stochastic sampling of a parent collective by offspring collectives. Other types of stochasticity, such as environmental stochasticity and measurement noise, are not considered and require future research.”

(5) Assumption about mutation rates

If setting the mutation rates to zero does not change the result of the simulations and the conclusion, what is the purpose of having the mutation rates \mu? Also, is the unidirectional (S -> F -> FF) mutation realistic? I didn't quite understand how the mutations could fit into the story of this paper.

This is a great point. We have added this to the beginning of Results to better motivate our study:

“We will start with a complete model where S mutates to F at a nonzero mutation rate µ. We made this choice because it is more challenging to attain or maintain the target frequency when the abundance of fast-growing F is further increased via mutations. This scenario is encountered in biotechnology: an engineered pathway will slow down growth, and breaking the pathway (and thus faster growth) is much easier than the other way around. When the mutation rate is set to zero, the same model can be used to capture collectives of two species with different growth rates.

See answer on common question 1.

(6) Minor points

In Figure 3b, it is not clear to me how the frequency difference for the Intra-collective and the Inter-collective selection is computed.

We added a description in caption 3b.

In Figure 5b, the gold region (success) near the FF is not visible. Maybe increase the size of the figure or have an inset for zoom-in. Why is the region not as big as the bottom gold region?

We increased the resolution of Fig 5b so that the gold region near FF is more visible.

We have added Fig 5c and the following explanation to the main text:

“From numerical simulations, we identified two accessible regions: a small region near FF and a band region spanning from S to F (gold in Fig. 5b i). Intuitively, the rate at which FF grows faster than S+F is greater than the rate at which F grows faster than S (see section VIII in Supplementary Information). Thus, the problem can initially be reduced to a two-population problem (i.e. FF versus F+S; Fig. 5c left), and then expanded to a three-population problem (Fig. 5c right).”

Recommendations For The Authors

Since the conclusion of the model greatly depends on the noise (variation) of F and S in the Gaussian distribution, it would be nice to have a plot where the y-axis is the variation in terms of frequency and the x-axis is the s_0 or f_0 (frequency). In the plot, I would love to see how the variation in the frequency depends on the initial frequency of S and F. Maybe this is just trivial.

In the SI, we added Fig6a, as per your request. Previous Fig6 became Fig6b.

Reviewer #2 (Public review):

The authors provide an analytical framework to model the artificial selection of the composition of communities composed of strains growing at different rates. Their approach takes into account the competition between the targeted selection at the level of the meta-community and the selection that automatically favors fast-growing cells within each replicate community. Their main finding is a tipping point or path-dependence effect, whereby compositions dominated by slow-growing types can only be reached by community-level selection if the community does not start and never crosses into a range of compositions dominated by fast growers during the dynamics.

These results seem to us both technically correct and interesting. We commend the authors on their efforts to make their work reproducible even when it comes to calculations via extensive appendices, though perhaps a table of contents and a short description of these appendices at the start of SI would help navigate them.

Thank you for the suggestion. We have added a paragraph at the beginning of SI.

The main limitation in the current form of the article is that it could clarify how its assumptions and findings differ from and improve upon the rest of the literature:

- Many studies discuss the interplay between community-level evolution and species- or strain-level evolution. But "evolution" can be a mix of various forces, including selection, drift/randomness, and mutation/innovation.

- This work's specificity is that it focuses strictly on constant community-level selection versus constant strain-level selection, all other forces being negligible (neither stochasticity nor innovation/mutation matter at either level, as we try to clarify now).

Note that intra-collective selection is not strictly “constant” in the sense that selection favoring F is the strongest at intermediate F frequency (Fig 3). However, we think that you mean that intra- and inter-collective selection are present in every cycle, and this is correct for our case, and for community selection in general.

- Regarding constant community-level selection, it is only briefly noted that "once a target frequency is achieved, inter-collective selection is always required to maintain that frequency due to the fitness difference between the two types" [pg. 3 {section sign}2]. In other words, action from the selector is required indefinitely to maintain the community in the desired state. This assumption is found in a fraction of the literature, but is still worth clarifying from the start as it can inform the practical applicability of the results.

This is a good point. We have added to abstract:

“Such collective selection is dictated by two opposing forces: during collective maturation, intra-collective selection acts like a waterfall, relentlessly driving the S-frequency to lower values, while during collective reproduction, inter-collective selection resembles a rafter striving to reach the target frequency. Due to this model structure, maintaining a target frequency requires the continued action of inter-collective selection.”

- More importantly, strain-level evolution also boils down here to pure selection with a constant target, which is less usual in the relevant literature. Here, (1) drift from limited population sizes is very small, with no meaningful counterbalancing of selection, (2) pure exponential regime with constant fitness, no interactions, no density- or frequency-dependence, (3) there is no innovation in the sense that available types are unchanging through time (no evolution of traits such as growth rate or interactions) and (4) all the results presented seem unchanged when mutation rate mu = 0 (as noted in Appendix III), meaning that the conclusions are not "about" mutation in any meaningful way.

With regard to point (1), Figure 4a (reproduced below) shows how Newborn size affects the region of achievable targets. Indeed at large Newborn size (e.g. 5000 and above), no target frequency is achievable (since drift is too small to generate sufficient inter-community variation and consequently all communities are dominated by fast-growing F). However at Newborn size of for example 1000, there are two regions of accessible target frequencies. At smaller Newborn size, all target frequencies become achievable due to drift becoming sufficiently strong.

With regard to points (2) and (3), we have added to Introduction

“To enable the derivation of an analytical expression, we have made the following simplifications.

First, growth is always exponential, without complications such as resource limitation, ecological interactions between the two populations, or density-dependent growth. Thus, the exponential growth equation can be used. Second, we consider only two populations (genotypes or species): the fast-growing F population with size F and the slow-growing S population with size S. We do not consider a spectrum of mutants or species, since with more than two populations, an analytical solution becomes very difficult.”

With regard to point (4), we view this as a strength rather than weakness. We have added the following to the beginning of Results and Discussions:

“We will start with a complete model where S mutates to F at a nonzero mutation rate µ. We made this choice because it is more challenging to attain or maintain the target frequency when the abundance of fast-growing F is further increased via mutations.”

“When the mutation rate is set to zero, the same model can be used to capture collectives of two species with different growth rates.”

See Point 1 of Common comments.

- Furthermore, the choice of mutation mechanism is peculiar, as it happens only from slow to fast grower: more commonly, one assumes random non-directional mutations, rather than purely directional ones from less fit to fitter (which is more of a "Lamarckian" idea). Given that mutation does not seem to matter here, this choice might create unnecessary opposition from some readers or could be considered as just one possibility among others.

We have added the following justification:

“This scenario is encountered in biotechnology: an engineered pathway will slow down growth, and breaking the pathway (and thus faster growth) is much easier than the other way around.”

It would be helpful to have all these points stated clearly so that it becomes easy to see where this article stands in an abundant literature and contributes to our understanding of multi-level evolution, and why it may have different conclusions or focus than others tackling very similar questions.

Finally, a microbial context is given to the study, but the assumptions and results are in no way truly tied to that context, so it should be clear that this is just for flavor.

We have deleted “microbial” from the title, and revised our abstract:

Recommendations For The Authors

(1) More details concerning our main remark above:

- The paragraph discussing refs [24, 33] is not very clear in how they most importantly differ from this study. Our impression is that the resource aspect is not very important for instance, and the main difference is that these other works assume that strains can change in their traits.

We are fairly sure that resource depletion is important in Rainey group’s study, as the attractor only evolved after both strains grew fast enough to deplete resources by the end of maturation. Indeed, evolution occurred in interaction coefficients which dictate the competition between strains for resources.

Regardless, you raised an excellent point. As discussed earlier, we have added the following:

“To enable the derivation of an analytical expression, we have made the following simplifications.

First, growth is always exponential, without complications such as resource limitation, ecological interactions between the two populations, or density-dependent growth. Thus, the exponential growth equation can be used. Second, we consider only two populations (genotypes or species): the fast-growing F population with size F and the slow-growing S population with size S. We do not consider a spectrum of mutants or species, since with more than two populations, an analytical solution becomes very difficult.”

- We would advise the main text to focus on mu = 0, and only say in discussion that results can be generalized.

Your suggestion is certainly good. However, given the large amount of work involved in a reorganisation, we have decided to adhere to our current narrative. However, as discussed earlier, we have added this at the beginning of Results to help orient readers:

“We will start with a complete model where S mutates to F at a nonzero mutation rate µ. We made this choice because it is more challenging to attain or maintain the target frequency when the abundance of fast-growing F is further increased via mutations.”

“When the mutation rate is set to zero, the same model can be used to capture collectives of two species with different growth rates.”

(2) We think the material on pg. 5 "Intra-collective evolution is the fastest at intermediate F frequencies, creating the "waterfall" phenomenon", although interesting, could be presented in a different way. The mathematical details on how to find the probability distribution of the maximum of independent random variables (including Equation 1) will probably be skipped by most of the readers (for experienced theoreticians, it is standard content; for experimentalists, it is not the most relevant), as such I would recommend displacing them to SM and report only the important results.

This is an excellent suggestion. We have put a sketch of our calculations in a box in the main text to help orient interested readers. As before, details are in SI.

Similarly, Equations 2, 3, and 4 are hard to read given the large amount of parameters and the low amount of simplification. Although exploring the effect of the different parameters through Figures 3 and 4 is useful, I think the role of the equations should be reconsidered:

i. Is it possible to rewrite them in terms of effective variables in a more concise way?

See Point 3 of Common comments.

ii. Is it possible to present extreme/particular cases in which they are easier to interpret?

We have focused on the case where the mutation rate is zero. This makes the mathematical expressions much simpler (see above).

(3) Is it possible to explain more in detail why the distribution of f_k+1 conditional to f_k^* is well approximated by a Gaussian? Also, have you explored to what extent the results would change if this were not true (in light of the few universal classes for the maximum of independent variables)?

Despite the appeal to the CLT and the histograms in the Appendix suggesting that the distribution looks a bit like a Gaussian at a certain scale, fluctuations on that scale are not necessarily what is relevant for the results - a rapid (and maybe wrong) attempt at a characteristic function calculation suggests that in your case, one does not obtain convergence to Gaussians unless we renormalize by S(t=0) and F(t=0), so it seems there is a justification missing in the text as is for the validity of this approximation (or that it is simply assumed).

See point 4 of Common comments.

Reviewer #3 (Public Reviews):

The authors address the process of community evolution under collective-level selection for a prescribed community composition. They mostly consider communities composed of two types that reproduce at different rates, and that can mutate one into the other. Due to such differences in 'fitness' and to the absence of density dependence, within-collective selection is expected to always favour the fastest grower, but the collective-level selection can oppose this tendency, to a certain extent at least. By approximating the stochastic within-generation dynamics and solving it analytically, the authors show that not only high frequencies of fast growers can be reproducibly achieved, aligned with their fitness advantage. Small target frequencies can also be maintained, provided that the initial proportion of fast growers is sufficiently small. In this regime, similar to the 'stochastic corrector' model, variation upon which selection acts is maintained by a combination of demographic stochasticity and of sampling at reproduction. These two regions of achievable target compositions are separated by a gap, encompassing intermediate frequencies that are only achievable when the bottleneck size is small enough or the number of communities is (disproportionately) larger.

A similar conclusion, that stochastic fluctuations can maintain the system over evolutionary time far from the prevalence of the faster-growing type, is then confirmed by analyzing a three-species community, suggesting that the qualitative conclusions of this study are generalizable to more complex communities.

I expect that these results will be of broad interest to the community of researchers who strive to improve community-level selection, but are often limited to numerical explorations, with prohibitive costs for a full characterization of the parameter space of such embedded populations. The realization that not all target collective functions can be as easily achieved and that they should be adapted to the initial conditions and the selection protocol is also a sobering message for designing concrete applications.

A major strength of this work is that the qualitative behaviour of the system is captured by an analytically solvable approximation so that the extent of the 'forbidden region' can be directly and generically related to the parameters of the selection protocol.

Thanks so much for these positive comments.

I however found the description of the results too succinct and I think that more could be done to unpack the mathematical results in a way that is understandable to a broader audience. Moreover, the phenomenon the authors characterize is of purely ecological nature. Here, mutations of the growth rate are, in my understanding, neither necessary (non-trivial equilibria can be maintained also when \mu =0) nor sufficient (community-level selection is necessary to keep the system far from the absorbing state) for the phenomenon described. Calling this dynamics community evolution reflects a widespread ambiguity, and is not ascribable just to this work. I find that here the authors have the opportunity to make their message clearer by focusing on the case where the 'mutation' rate \mu vanishes (Equations 39 & 40 of the SI) - which is more easily interpretable, at least in some limits - while they may leave the more general equations 3 & 4 in the SI.

See points 1-4 of Common comments.

Combined with an analysis of the deterministic equations, that capture the possibility of maintaining high frequencies of fast growers, the authors could elucidate the dynamics that are induced by the presence of a second level of selection, and speculate on what would be the result of real open-ended evolution (not encompassed by the simple 'switch mutations' generally considered in evolutionary game theory), for instance discussing the invasibility (or not) of mutant types with slightly different growth rates.

Indeed, evolution is not restricted to two types. However, our main goal here is to derive an analytical expression, and it was difficult for even two types. For three-type collectives, we had to resort to simulations. Investigating the case where fitness effects of mutations are continuously distributed is beyond the scope of this study.

The single most important model hypothesis that I would have liked to be discussed further is that the two types do not interact. Species interactions are not only essential to achieve inheritance of composition in the course of evolution but are generally expected to play a key role even on ecological time scales. I hope the authors plan to look at this in future work.

In our system, the S and F do interact in a competitive fashion: even though S and F are not competing for nutrients (which are always in excess), they are competing for space. This is because a fixed number of cells are transferred to the next cycle. Thus, the presence of F will for example reduce the chance of S being propagated. We have added this clarification to our main text:

“Note that even though S and F do not compete for nutrients, they compete for space: because the total number of cells transferred to the next cycle is fixed, an overabundance of one population will reduce the likelihood of the other being propagated.”

Recommendations For The Authors

I felt the authors could put some additional effort into making their theoretical results meaningful for a population of readers who, though not as highly mathematically educated as they are, can nonetheless appreciate the implications of simple relations or scaling. Below, you find some suggestions:

(1) In order to make it clear that there is a 'natural' high-frequency equilibrium that can be reached even in the absence of selection, the authors could examine first the dynamics of the deterministic system in the absence of mutations, and use its equilibria to elucidate the combined role of the 'fitness' difference \omega and of the generation duration \tau in setting its value. The fact that these parameters always occur in combination (when there are no mutations) is a general and notable feature of the stochastic model as well. Moreover, this model would justify why you only focus on decreasing the frequency in the new generation.

Note that the ‘natural’ high-frequency equilibrium in the absence of collective selection is when fast grower F becomes fixed in the population. Following your suggestion, we have introduced two parameters 𝑅τ and 𝑊τ to reflect the coupling between ‘fitness’ and ‘generation duration’:

(2) Since the phenomenon described in the paper is essentially ecological in nature (as the author states, it does not change significantly if the 'mutation rate' \mu is set to zero), I would put in the main text Equations 39 & 40 of the SI in order to improve intelligibility.

See Point 2 at the beginning of this letter.

These equations can be discussed in some detail, especially in the limit of small f^*_k, where I think it is worth discussing the different dependence of the mean and the variance of the frequency distribution on the system's parameters.

This is a great suggestion. We have added the following:

“In the limit of small , Equation (3) becomes f while Equation (4) becomes N0) and fold-change in F/S during maturation (. Thus, both Newborn size (Wτ) are important determinants of selection progress.

(3) I would have appreciated an explanation in words of what are the main conceptual steps involved in attaining Equation 2, the underlying hypotheses (notably on community size and distributions), and the expected limits of validity of the approximation.

See points 3 and 4 at the beginning of this letter.

(4) I think that some care needs to be put into explaining where extreme value statistics is used, and why is the median of the conditional distribution the most appropriate statistics to look at for characterizing the evolutionary trajectory (which seems to me mostly reliant on extreme values).

Great point! We added an explanation of using median value in Box 1.

and also added figure 7 to explaining it in SI.

Showing in a figure the different distributions you are considering (for instance, plotting the conditional distribution for one generation in the trajectories displayed in Figure 2) would be useful to understand what information \bar f provides on a sequence of collective generations, where in principle there may be memory effects.

Thanks for this suggestion. We have added to Fig 2d panel to illustrate the shape and position of F frequency distributions in each step in the first two selection cycles.

(5) Similarly, I do not understand why selecting the 5% best communities should push the system's evolution towards the high-frequency solution, instead of just slowing down the improvement (unless you are considering the average composition of the top best communities - which should be justified). I think that such sensitivity to the selection intensity should be appropriately referenced and discussed in the main text, as it is a parameter that experimenters are naturally led to manipulate.

In the main text, we have added this explanation:

“In contrast with findings from an earlier study [23], choosing top 1 is more effective than the less stringent “choosing top 5%”. In the earlier study, variation in the collective trait is partly due to nonheritable factors such as random fluctuations in Newborn biomass. In that context, a less stringent selection criterion proved more effective, as it helped retain collectives with favorable genotypes that might have exhibited suboptimal collective traits due to unfavorable nonheritable factors. However, since this study excludes nonheritable variations in collective traits, selecting the top 1 collective is more effective than selecting the top 5% (see Fig. 11 in Supplementary Information).”

(6) Equation 1 could be explained in simpler terms as the product between the probability that one collective reaches the transmitted value times the probability that all others do worse than that. The current formulation is unclear, perhaps just a matter of English formulation.

We have revised our description to state:

“Equation (1) can be described as the product between two terms related to probability: (i) describes the probability density that any one of the g Adult collectives achieves f given , and (ii) f and thus not selected.” describes the probability that all other g – 1 collectives achieve frequencies above

(7) I think that the discussion of the dependence of the boundaries of the 'waterfall' region with the difference in growth rate \omega is important and missing, especially if one wants to consider open-ended evolution of the growth rate - which can occur at steps of different magnitude.

We added a new chapter and figure in supplementary information on the threshold values when \omega varies. As expected, smaller \omega enlarges the success area.

We have also added a new figure panel to show how maturation time affects selection efficacy.

(8) Notations are a bit confusing and could be improved. First of all, in most equations in the main text and SI, what is initially introduced as \omega appears as s. This is confusing because the letter s is also used for the frequency of the slow type.

The letter S is used to denote an attribute of cells (S cells), the type of cells (Equations 1-3 of the SI) and the number of these cells in the population, sometimes with different meanings in the same sentence. This is confusing, and I suggest referring to slow cells or fast cells instead (or at least to S-cells and F-cells), and keeping S and F as variables for the number of cells of the two types.

All typos related to the notation have been fixed. We use S and F as types, and S and F (italic) and population numbers.

(9) On page 3, when introducing the sampling of newborns as ruled by a binomial distribution, the information that you are just transmitting one collective is needed, while it is conveyed later.

We have added this emphasis:

“At the end of a cycle, a single Adult with the highest function (with F frequency f closest to the target frequency ) is chosen to reproduce g Newborn collectives each with N0 cells (‘Selection’ and ’Reproduction’ in Fig. 1).”

(10) I found that the abstract talks too early about the 'waterfall' phenomenon. As this is a concept introduced here, I suggest the authors first explain what it is, then use the term. It is a useful metaphor, but it should not obscure the more formal achievements of the paper.

We feel that the “waterfall” analogy offers a gentle helping hand to orient those who have not thought much about the phenomenon. We view abstract as an opportunity to attract readership, and thus the more accessible the better.

(11) In the SI there are numerous typos and English language issues. I suggest the authors read carefully through it, and add line numbers to the next version so that more detailed feedback is possible.

Thank you for going through SI. We have gone through the SI, and fixed problems.

Version 2

Abstract

Collectives, such as microbial communities, can perform functions beyond the capability of individual members. Enhancing these collective functions through artificial selection, however, presents significant challenges. Here, we explore the ‘rafting-a-waterfall’ phenomenon, a metaphor illustrating how the success in achieving a target population composition in microbial collectives depends on both the target characteristics and initial conditions. Specifically, collectives comprising fast-growing (F) and slow-growing (S) individuals were grown for a period of “maturation” time, and the collective with S-frequency closest to the target value is chosen to “reproduce” (inoculate) offspring collectives. Such collective selection is dictated by two opposing forces: during collective maturation, intra-collective selection acts like a waterfall, relentlessly driving the S-frequency to lower values, while during collective reproduction, inter-collective selection resembles a rafter striving to reach the target frequency. Due to this model structure, maintaining a target frequency requires the continued action of inter-collective selection. Using simulations and analytical calculations, we show that intermediate target S frequencies are the most challenging, akin to a target within the vertical drop of a waterfall, rather than above or below it. This arises because intra-collective selection is the strongest at intermediate S-frequencies, which can overpower inter-collective selection. While achieving low target S frequencies is consistently feasible, attaining high target S-frequencies requires an initially high S-frequency — much like a raft that can descend but not ascend a waterfall. The range of attainable target frequencies depends on the initial population size of the collectives: as the population size in Newborn collectives increases, the region of achievable target frequency is reduced until no frequency is achievable. In contrast, the number of collectives under selection plays a less critical role. In scenarios involving more than two populations, the evolutionary trajectory must navigate entirely away from the metaphorical ‘waterfall drop.’ Our findings illustrate that the strength of intra-collective evolution is frequency-dependent, with implications in experimental planning.

Introduction

Microbial collectives can carry out functions that arise from interactions among member species. These functions, such as waste degradation [1, 2], probiotics [3], and vitamin production [4], can be useful for human health and biotechnology. To improve collective functions, one can perform artificial selection (directed evolution) on collectives: Low-density “Newborn” collectives are allowed to “mature” during which cells proliferate and possibly mutate, and community function develops. “Adult” collectives with high functions are then chosen to reproduce, each seeding multiple offspring Newborns. Artificial selection of collectives have been attempted both in experiments [5–19] and in simulations [20–31], often with unimpressive outcomes.

One of the major challenges in selecting collectives is to ensure the inheritance of a collective function [32, 33]. Inheritance from a parent collective to offspring collectives can be compromised by changes in genotype and species compositions. During maturation of a collective, genotype compositions within each species can change due to intra-collective selection favoring fast-growing individuals, while species compositions can change due to ecological interactions. Furthermore, during reproduction of a collective, genotype and species compositions of offspring can vary stochastically from those of the parent.

Here, we consider the selection of collectives comprising two or three populations with different growth rates, and our goal is to achieve a target composition in the Adult collective. This is a common quest: whenever a collective function depends on both populations, collective function is maximised, by definition, at an intermediate frequency (e.g. too little of either population will hamper function [23]). A selection cycle consists of three stages (Fig. 1). During collective maturation, intra-collective selection favors fast-growing individuals within a collective. At the end of maturation, inter-collective selection acts on collectives and favors those achieving the target composition. Finally during collective reproduction, offspring collectives sample stochastically from the parents, a process dominated by genetic drift.

We mathematically examine the selection of composition in collectives consisting of populations growing at different rates. Earlier work has demonstrated that nearly any target species composition can be achieved when selecting communities of two competing species with unequal growth rates [24, 34], so long as the shared resource is depleted during collective maturation [24]. In this case initially, both species evolved to grow faster, and the slower-growing species was preserved due to stochastic fluctuations in species composition during collective reproduction. Eventually, both species evolved to grow sufficiently fast to deplete the shared resource during collective maturation, and evolution in competition coefficients then acted to stabilize species ratio to the target value [24]. Regardless, earlier studies are often limited to numerical explorations, with prohibitive costs for a full characterization of the parameter space for such nested populations (population of collectives, and populations of mutants within a collective). We made simplifying assumptions so that we can analytically examine the evolutionary tipping point between intra-collective and inter-collective selection. We show that this tipping point creates a “waterfall” effect which restricts not only which target compositions are achievable, but also the initial composition required to achieve the target. We also investigate how the range of achievable target composition is affected by the total population size in Newborns and the total number of collectives under selection. Finally, we show that the waterfall phenomenon extends to systems with more than 2 populations.

Schematic for artificial selection on collectives.

Each selection cycle begins with a total of g Newborn collectives, each with N0 total cells of slow-growing S population (light gray dots) and fast-growing F population (dark gray dots). During maturation (over time τ), S and F cells divide at rates rS and rS + ω (ω > 0), respectively, and S mutates to F at rate µ. In the selection stage, the Adult collective with F frequency f closest to the target composition is chosen to reproduce g Newborns for the next cycle. Newborns are sampled from the chosen Adult (yellow star) with N0 cells per Newborn. The selection cycle is then repeated until the F frequency reaches a steady state, which may or may not be the target composition. To denote a variable x of i-th collective in cycle k at time t (0 ≤ t ≤ τ), we use notation where x ∈ {S, F, s, f}. Note that time t = 0 is for Newborns and t = τ is for Adults.

Results and discussions

To enable the derivation of an analytical expression, we have made the following simplifications. First, growth is always exponential, without complications such as resource limitation, ecological interactions between the two populations, or density-dependent growth. Thus, the exponential growth equation can be used. Second, we consider only two populations (genotypes or species): the fast-growing F population with size F and the slow-growing S population with size S. We do not consider a spectrum of mutants or species, since with more than two populations, an analytical solution becomes very difficult. Finally, the single top-functioning community is chosen to reproduce, which allows us to employ the simplest version of the extreme value theory (see section below for further justification.)

Our goal is to select for collective composition in terms of F frequency f = F/(S + F), or equivalently, S frequency s = 1 − f. More precisely, we want collectives such that after maturation time τ, f (τ) is as close to the target value as possible (Fig. 1). Note that even if the target frequency has been achieved, since F frequency will always increase during maturation, inter-collective selection is required in each cycle to maintain the target frequency.

We will start with a complete model where S mutates to F at a nonzero mutation rate µ. We made this choice because it is more challenging to attain or maintain the target frequency when the abundance of fast-growing F is further increased via mutations. This scenario is encountered in biotechnology: an engineered pathway will slow down growth, and breaking the pathway (and thus faster growth) is much easier than the other way around. When the mutation rate is set to zero, the same model can be used to capture collectives of two species with different growth rates. We show that intermediate F frequencies or equivalently, intermediate S frequencies, are the hardest targets to achieve. We then show that similar conclusions hold when selecting for a target composition in collectives of more than two populations.

Model structure

A selection cycle (Fig. 1) starts with a total of g Newborn collectives. At the beginning of cycle k (t = 0), each Newborn collective has a fixed total cell number where and, denote the numbers of S and F cells in collective i (1 ≤ i ≤ g) at time t (0 ≤ t ≤ τ) of cycle k. The average F frequency among the g Newborn collectives is fk,0, such that the initial F cell number in each Newborn is drawn from the binomial distribution Binom .

Collectives are allowed to grow for time τ (‘Maturation’ in Fig. 1). During maturation, S and F grow at rates rS and rS + ω (ω > 0), respectively. If maturation time τ is too small, a matured collective (“Adult”) does not have enough cells to reproduce g Newborn collectives with N0 cells. On the other hand, if maturation time τ is too long, fast-growing F will take over. Hence we set the maturation time τ = ln(g +1)/rS, which guarantees sufficient cells to produce g Newborn collectives from a single Adult collective. At the end of a cycle, a single Adult with the highest function (with F frequency f closest to the target frequency ) is chosen to reproduce g Newborn collectives each with N0 cells (‘Selection’ and ‘Reproduction’ in Fig. 1). Note that even though S and F do not compete for nutrients, they compete for space: because the total number of cells transferred to the next cycle is fixed, an overabundance of one population will reduce the likelihood of the other being propagated.

Collective function is dictated by the Adult’s F frequency f. Among all Adult collectives, the selected Adult is the one whose F frequency is closest to the target value, . In contrast with findings from an earlier study [23], choosing top 1 is more effective than the less stringent “choosing top 5%”. In the earlier study, variation in the collective trait is partly due to nonheritable factors such as random fluctuations in Newborn biomass. In that context, a less stringent selection criterion proved more effective, as it helped retain collectives with favorable genotypes that might have exhibited suboptimal collective traits due to unfavorable nonheritable factors. However, since this study excludes nonheritable variations in collective traits, selecting the top 1 collective is more effective than selecting the top 5% (see Fig. 11 in Supplementary Information).

The selected Adult, with F frequency denoted as f *, is then used to reproduce g offspring collectives, each with N0 total cells. The number of F cells in a Newborn follows a binomial distribution B(N0, f *). By repeating the selection cycle, we aim to achieve and maintain the target composition .

Overall our model considers mutational stochasticity, as well as demographic stochasticity in terms of stochastic birth and stochastic sampling of a parent collective by offspring collectives. Other types of stochasticity, such as environmental stochasticity and measurement noise, are not considered and require future research.

The success of collective selection is constrained by the target composition, and sometimes also by the initial composition

Since intra-collective selection favors F, we expect that a higher target (a lower target ŝ) is easier to achieve. By 'achieve', we mean the absolute error d between the target frequency and the selected frequency averaged among independent simulations (f *) is smaller than 0.05 (i.e.).

We fixed N0, the total population size of a Newborn, to 1000, and obtained selection dynamics for various initial and target F frequencies by implementing stochastic simulations (Sec. I in Supplementary Information). If the target is high (e.g. 0.9, Fig. 2a magenta), selection is successful (computed absolute errors in Supplementary Information Fig. 4): regardless of the initial frequency, f * of the chosen collective eventually converges to the target and stays around it. In contrast, without collective-level selection (e.g. choosing a random collective to reproduce), F frequency increases until F reaches fixation (Supplementary information Fig. 3b).

Initial and target compositions determine the success of artificial selection on collectives.

(a-c) F frequency of the selected Adult collective (f *) over cycles at different target values (dashed lines). between f L and f H (orange dotted and solid line segments) is inaccessible where selection will fail. a A high target F frequency (e.g. ; magenta dotted line) can be achieved from any initial frequency (black dots). b An intermediate target frequency (e.g. ; green dotted line) is never achievable, as all initial conditions converge to near f H. c A low target frequency (e.g.) is achievable, but only from initial frequencies below f L. For initial frequencies at f L, stochastic outcomes (grey curves) are observed: while some replicates reached the target frequency, others reached f H. For parameters, we used S growth rate rS = 0.5, F growth advantage ω = 0.03, mutation rate µ = 0.0001, maturation time τ≈ 4.8, and N0 = 1000. The number of collectives g = 10. Each black line is averaged from independent 300 realizations. d Inter-collective selection opposes intra-collective selection. We plot probability density distributions of F frequency f during two consecutive cycles when selection is successful. Data correspond to cycles 31 and 32 from the second lowest initial point in c. Δf is selection progress (see Box). Black triangle: median. e Two accessible regions (gold). Either high ; region 2) or low starting from low initial f (and ; region 1) can be achieved. We theoretically predict (by numerically integrating Eq. (1)) f H (orange solid line) and f L (orange dotted line), which agree with simulation results (gold regions). f Example trajectories from initial compositions (black dots) to the target compositions (dashed lines). The gold areas indicate the region of initial frequencies where the target frequency can be achieved. g The tension between intra-collective selection and inter-collective selection creates a “waterfall” phenomenon. See the main text for details.

In contrast, an intermediate target frequency (e.g. ; Fig. 2b green) is never achievable. High initial F frequencies (e.g. 0.6 and 0.95) decline toward the target, but stabilize at the “higher-threshold” f H (∼ 0.7, solid orange line segment in Fig. 2a-c) above the target. Low initial F frequencies (e.g. 0) increase toward the target, but then overshoot and stabilize at the f H value.

If the target frequency is low (e.g.; Fig. 2c dark blue), artificial selection succeeds when the initial frequency is below the “lower-threshold” f L (dotted orange line segment in Fig. 2a-c). Initial F frequencies above f L (e.g. 0.45 and 0.95) converge to f H instead. Initial F frequencies near f L display stochastic trajectories, converging to either f H or f L.

To achieve target , inter-collective selection must overcome intra-collective selection. We can visualize the distributions of f over two consecutive cycles (bottom to top, Fig. 2d) where f started above target . When Newborns matured into Adults, the distribution of f up-shifted due to intra-collective selection. The distribution of f was then down-shifted toward the target due to inter-collective selection. If the magnitude of down-shift exceeded that of up-shift, progress toward the target was made. During reproduction, the distribution of f retained the same mean but became broader due to stochastic sampling by the Newborns from their parent.

In summary, two regions of target frequencies are “accessible” in the sense that a target value in the region can be reached (gold in Fig. 2e, f): (1) target frequencies above or (2) target frequencies below and starting at an average frequency below .

Intra-collective evolution is the fastest at intermediate F frequencies, creating the “waterfall” phenomenon

To understand what gives rise to the two accessible regions, we calculated Δf, the selection progress in F frequency over two consecutive cycles (Box 1, Eq. (2)). The solution (Fig. 3a, green) has the same shape as results from numerically integrating Eq. (1) (Fig. 3a, orange) and from stochastic simulations (Fig. 3a, blue).

Intra-collective selection and inter-collective selection jointly set the boundaries for selection success.

a The change in F frequency over one cycle. When is sufficiently low or high, inter-collective selection can lower the F frequency to below . The points where Δf = 0 (in the orange line) are denoted as f L and f H, corresponding to the boundaries in Fig. 2. b The distributions of frequency differences obtained by 1000 numerical simulations. The cyan, purple, and black box plots respectively indicate the changes in F frequency after intra-collective selection (the mean frequency among the 100 Adults minus the mean frequency among the 100 Newborns during maturation), after inter-collective selection (the frequency of the 1 selected Adult minus the mean frequency among the 100 Adults), and over one selection cycle (the frequency of the selected Adult of one cycle minus that of the previous cycle). The box ranges from 25% to 75% of the distribution, and the median is indicated by a line across the box. The upper and lower whiskers indicate maximum and minimum values of the distribution. ***: P < 0.001 in an unpaired t-test.

If Δf is negative, then inter-collective selection will succeed in countering intra-collective selection and reducing f toward the target. Δf is negative if the selected is low or high, but not if it is intermediate between f L and f H (Fig. 3a). This is because f increase during maturation is the most drastic when Newborn f is intermediate (Fig. 3b), for intuitive reasons: when Newborn f is low, the increase in f will be minor; when Newborn f is high, the fitness advantage of F over the population average is small and hence the increase is also minor. Thus, when Newborn F frequency is intermediate, intra-collective selection is the strongest and may overwhelm inter-collective selection (Fig. 3b and Fig. 6a in Supplementary Information). Not surprisingly, similar conclusions are derived where S and F are slow-growing and fast-growing species which cannot be converted through mutations (Sec. IV and Fig. 8 in Supplementary Information).

Thus, inter-collective selection is akin to a raftman rowing the raft to a target, while intra-collective selection is akin to a waterfall. This metaphor is best understood in terms of S frequency s = 1 − f. The lower-threshold f L corresponds to higher-threshold in sH = 1 − f L, while higher-threshold f H corresponds to lower-threshold in sL = 1 − f H. Intra-collective selection is akin to a waterfall, driving the S frequency s from high to low (Fig. 2g). Intra-collective selection acts the strongest when s is intermediate (sL < s < sH), similar to the vertical drop of the fall. Intra-collective selection acts weakly at high (> sH) or low (< sL) s, similar to the gentle sloped upper and lower pools of the fall (regions 1 and 2 of Fig. 2e, g). Thus, an intermediate target frequency can be impossible to achieve − a raft starting from the upper pool will be flushed down to sL (f H), while a raft starting from the lower pool cannot go beyond sL (f H). In contrast, a low target S frequency (in the lower pool) is always achievable. Finally, a high target S frequency (in the upper pool) can only be achieved if starting from the upper pool (as the raft can not jump to the upper pool if starting from below).

Changes in the distribution of F frequency f after one cycle

We consider the case where , the F frequency of the selected Adult at cycle k, is above the target value . This case is particularly challenging because intra-collective evolution favors fast-growing F and thus will further increase f away from the target. From , Newborns of cycle k + 1 will have f fluctuating around , and after they mature, the minimum f is selected . If the selected composition at cycle k + 1 can be reduced compared to that of cycle k (i.e. ), the system can evolve to the lower target value.

To find values such that , we used the median value of the conditional probability distribution Ψ of (the F frequency of the selected Adult at cycle k + 1 being f) given the selected at cycle k (mathematical details in Sec. II in Supplementary Information). If the median value (Median) is smaller than , then selection will likely be successful since the selected Adult in cycle k + 1 has more than 50% chance to have a reduced F frequency compared to cycle k.

There are two points where the median values are the same as (Fig. 3a), which are assigned as lower-threshold (f L) and higher-threshold (f H).

Following the extreme value theory, the conditional probability density function is Equation (1) can be described as the product between two terms related to probability: (i) describes the probability density that any one of the g Adult collectives achieves f given , and (ii) describes the probability that all other g − 1 collectives achieve frequencies above f and thus not selected.

Since computing the exact formula of Adults’ f distribution in cycle k + 1 is hard, we approximate it as Gaussian with mean and variance .The Gaussian approximation on Eq. (1) requires sharp Gaussian distributions of S(τ) and F (τ) (i.e. and ). Compared to Gaussian, the exact S(τ) (negative binomial) distribution and F (τ) (Luria-Delbrück) distribution are right-skewed and heavy-tailed. However these problems are alleviated when the initial numbers of S and F cells are not small (on the order of 100). Additionally, the sharpness of distributions could be achieved (see Fig. 1 in Supplementary Information).

To obtain an analytical solution of the change in f over one cycle, we first assume that in a Newborn collective, the number of S cells is distributed as Gaussian with mean and variance . Then, the number of F cells, F0 = N0 − S0, is distributed as Gaussian with mean and variance . From these, we can calculate for Adult collectives the mean and variance of population sizes F (τ) (i.e. ) and S(τ) (i.e.) (mathematical details in Sec. I in Supplementary Information). This task is simplified by the exponential growth of S and describes the fold-growth of S over maturation time τ, and since ω is the fitness advantage of F over S, Wτ = eωτ describes the fold change of F/S over time τ. From (mutation rate scaled with the fitness difference), (F frequency in the selected collective at cycle k), N0 (Newborn size), (relative fitness advantage), we can calculate the mean and variance of F frequency among the Adults of k + 1 cycle , detailed formula in Eqs. (40) and (41) in Supplementary Information).

Selection progress - the difference between the median value of the conditional probability distribution and the selected frequency of (Sec. II in Supplementary Information) - can be expressed as: where Φ−1(…) is the inverse cumulative function of standard normal distribution (see main text for an example). We chose median because compared to mean, it is easier to get analytical expression since Φ−1(…) is known in a closed form. Regardless, using median generated results similar to simulations (Fig. 7 in Supplementary Information). As expected, selection progress Δf is governed by both the mean and the variation (σf (τ)) in f among Adults.

When mutation rate and σf (τ) can be simplified to: and In the limit of small , Equation (3) becomes while Equation (4) becomes . Thus, both Newborn size (N0) and fold-change in F/S during maturation (Wτ) are important determinants of selection progress.

Manipulating experimental setups to expand the achievable target region

In Eq. (2) (Box 1), selection progress Δf depends on the total number of collectives under selection (g). Δf also depends on the mean and the standard deviation of Adult F frequency and σf (τ). Eqs. (3) and (4) (Box 1) provide simplified expressions of and σf (τ) when mutation rate µ has been set to 0. When mutation rate µ is not zero (Eqs. (40) and (41) in Section II of Supplementary Information), selection progress is additionally influenced by (mutation rate µ scaled with fitness difference ω).

Our goal is to make Δf as negative as possible so that any increase in f during collective maturation may be reduced. From Eq. (2) (Box 1), a small f (τ) will facilitate collective-level selection. Additionally, a large σf (τ) will also facilitate collective-level selection due to negative . Note that since for corresponding to the number y such that the probability of a standard normal random variable being less than or equal to y is is negative.

From Eq. (4) (Box 1), σf (τ) will be large if Newborn size N0 is small. Indeed, as Newborn size N0 declines, the region of achievable target frequency expands (larger gold area in Fig. 4a). If the Newborn size N0 is sufficiently small (e.g. ≤ 700 in our parameter regime), any target frequency can be reached. An analytical approximation of the maximal Newborn size permissible for all target frequencies is given in section III in Supplementary Information.

Expanding the success region for artificial collective selection.

a Reducing the population size in Newborn N0 expands the region of success. In gold area, the probability that becomes smaller than in a cycle is more than 50%. We used g = 10 and τ ≈ 4.8. Figures 2-3 correspond to N0 = 1000. Black dotted line indicates the critical Newborn size , below which all target frequencies can be achieved. b Increasing the total number of collectives g also expands the region of success, although only slightly. We used a fixed Newborn size N0 = 1000. The maturation time τ (τ = log(100)/rS≈ 9.2) is set to be long enough so that an Adult can generate at least 100 Newborns. c Increasing the maturation time τ shrinks the region of success. We used a fixed Newborn size N0 = 1000 and number of collectives g = 10.

From Eqs. (3) and (4) (Box 1), maturation time τ affects through (the fold change in F/S over τ), and affects additionally through (fold-growth of S over τ). Longer τ increases and is thus detrimental to selection progress. The relationship between σf (τ) and τ is not monotonic (Fig. 6c in Supplementary Information), meaning that an intermediate value of τ is the best for achieving large σf (τ). However, the effect of dominates that of σf (τ) and therefore, the region of success monotonically reduces with longer maturation time (Fig. 4c). Similarly, will be small if ω (fitness advantage of F over S) is small. Indeed, as ω become larger, the region of success becomes smaller (Fig. 9 in Supplementary Information).

g, the number of collectives under selection, also affects selection outcomes. As g increases, the value of becomes more negative, and so does Δf - meaning collective-level selection will be more effective. Intuitively, with more collectives, the chance of finding a f closer to the target is more likely. Thus, a larger number of collectives broadens the region of success (Fig. 4b). However, the effect of g is not dramatic. To see why, we note that the only place that g appears is Eq. (2) in . When g becomes large, is asymptotically expressed as ln (Section II in Supplementary Information)[35], and thus does not change dramatically as g varies.

The waterfall phenomenon in a higher dimension

To examine the waterfall effect in a higher dimension, we investigate a three-population system where a faster-growing population (FF) grows faster than the fast-growing population (F) which grows faster than the slow-growing population (S) (Fig. 5a and Sec. VIII in Supplementary Information). In the three-population case, the evolutionary trajectory travels in a two-dimensional plane. A target population composition can be achieved if inter-collective selection can sufficiently reduce the frequencies of F as well as FF (accessible regions, gold in Fig. 5b).

In higher dimensions, the success of artificial selection requires the entire evolutionary trajectory remaining in the accessible region.

a During collective maturation, a slow-growing population (S) (with growth rate rS ; light gray) can mutate to a fast-growing population (F) (with growth rate rS + ω; medium gray), which can mutate further into a faster-growing population (FF) (with growth rate rS + 2ω; dark gray). Here, the rates of both mutational steps are µ, and ω > 0. b Evolutionary trajectories from various initial compositions (open circles) to various targets. Intra-collective evolution favors FF over F (vertical blue arrow) over S (horizontal blue arrow). The accessible regions are marked gold (see Sec. I in Supplementary Information). We obtain final compositions starting from several initial compositions while aiming for different target compositions in i, ii, and iii. The evolutionary trajectories are shown in dots with color gradients from the initial to final time. (i) A target composition with a high FF frequency is always achievable. (ii) A target composition with intermediate FF frequency is never achievable. (iii) A target composition with low FF frequency is achievable only if starting from an appropriate initial composition such that the entire trajectory never meanders away from the accessible region. The figures are drawn using mpltern package [36]. c The accessible region in the three-population problem is interpreted as an extension of the two-population problem. First the accessible region between FF and S+F is given, and then the S+F region is stretched into S and F.

From numerical simulations, we identified two accessible regions: a small region near FF and a band region spanning from S to F (gold in Fig. 5b i). Intuitively, the rate at which FF grows faster than S+F is greater than the rate at which F grows faster than S (see section VIII in Supplementary Information). Thus, the problem can initially be reduced to a two-population problem (i.e. FF versus F+S; Fig. 5c left), and then expanded to a three-population problem (Fig. 5c right).

Similar to the two-population case, targets in the inaccessible region are never achievable (Fig. 5b ii), while those in the FF region are always achievable (Fig. 5b i). Strikingly, a target composition in an accessible region may not be achievable even when the initial composition is within the same region: once the composition escapes the accessible region, the trajectory cannot return back to the accessible region (Fig. 5biii, the leftmost initial condition). However, if the initial position is closer to the target in the accessible region, the target becomes achievable (Fig. 5b iii, initial condition near the bottom). Note that here, selection outcome is path-dependent in the sense of being sensitive to initial conditions. This phenomenon is distinct from hysteresis where path-dependence results from whether a tuning parameter is increased or decreased.

In conclusion, we have investigated the evolutionary trajectories of population compositions in collectives under selection, which are governed by intra-collective selection (which favors fast-growing populations) and inter-collective selection (which, in our case, strives to counter fast-growing populations). Intra-collective selection has the strongest effect at intermediate frequencies of faster-growing populations, potentially creating an inaccessible region of target frequency, analogous to the vertical drop of a waterfall. High and low target frequencies are both accessible, analogous to the lower- and the upper-pools of a waterfall, respectively. A less challenging target (high ; low ŝ) is achievable from any initial position. In contrast, a more challenging target (low ; high ŝ) is only achievable if the entire trajectory is contained within the region, similar to a raft striving to reach a point in the upper-pool must start at and remain in the upper pool. Our work suggests that the strength of intra-collective selection is not constant, and that strategically choosing an appropriate starting point can be essential for successful collective selection.

Methods

Stochastic simulations

A selection cycle is composed of three steps: maturation, selection, and reproduction. At the beginning of the cycle k, a collective i has slow-growing cells and fast-growing cells. At the first cycle, mean F frequency of collectives are set to be is sampled from the binomial distribution with mean . Then, cells are in the collective i. In the maturation step, we calculate and by using stochastic simulation. We can simulate the division and mutation of each individual cell stochastically by using tau-leaping algorithm [37, 38](see Fig. 3a and b in Supplementary Information). However, individual-based simulations require a long computing time. Instead, we randomly sample and from the joint probability density distribution . To obtain , we solve the master equation which describes a time evolution of probability distribution under the random processes (see Sec. I in Supplementary Information). We assumed that and are independent (as S and F populations grow independently without ecological interactions), and thus is product of two probability density functions and . Each distribution follows a Gaussian distribution, with the mean and variance numerically obtained from ordinary differential equations derived from the master equation (see Sec. I in Supplementary Information). We choose the collective with the closest frequency to the target to generates g Newborns. The number of F cells is sampled from the binomial distribution with the mean of . We start a new cycle with those Newborn collectives. Then, the number of S cells in a collective i is .

Analytical approach to the conditional probability

The conditional probability distribution of observing at a given is calculated by the following procedure. Given the selected collective in cycle k with , the collective-level reproduction proceeds by sampling g Newborn collectives with N0 cells in cycle k + 1. Each Newborn collective contains certain F numbers at the beginning of the cycle k + 1, which can be mapped into with the constraint of N0 cells. If the number of cells in the selected collective is large enough, the joint conditional distribution function is well described by the product of g independent and identical Gaussian distribution 𝒩 (µ, σ2). So we consider the frequencies of g Newborn collectives as g identical copies of the Gaussian random variable fk+1,0. The mean and variance of fk+1,0 are given by and . Then, the conditional probability distribution function of fk+1,0 being ζ is given by After the reproduction step, the Newborn collectives grow for time τ. The frequency is changed from the given frequency ζ to f by division and mutation processes. We assume that the frequency f of an Adult is also approximated by a Gaussian random variable . The mean and variance are calculated by using means and variances of S and F (see Sec. II in Supplementary Information.) Since and also depend on ζ, the conditional probability distribution function of fk+1,τ being f is given by The conditional probability distribution of an Adult collective in cycle k + 1 (fk+1,τ) to have frequency f at a given is calculated by multiplying two Gaussian distribution functions and integrating overall ζ values, which is given by Since we select the minimum frequency among g identical copies of fk+1,τ, the conditional probability distribution function of follows a minimum value distribution, which is given in Eq. (1). Here, for the case of , the selected frequency is the minimum frequency . So we have by replacing with .

We assume that the conditional probability distribution in Eq. (7) follows a normal distribution, whose mean and variances are describe by Eq. (40) and Eq. (41). Then, the extreme value theory [39] estimates the median of the selected Adult by The selection progress Δf in Eq. (2) is obtained by subtracting from Eq. (8).